Complete Real-Time Liver Model including Glisson Capsule, Vascular Network and Parenchyma

### 2013-02-28 Doc/Miccai2013/miccai2013.tex
bibliography, comments, figures for aspiration
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@530 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

%\Figure{width}{name}{caption}
\newcommand{\Figure}[3]{%
\begin{figure}[htb]
  \centering
  \includegraphics[width=#1]{#2}
  \caption{\label{fig-#2}#3}
\end{figure}}

@@

In case of human liver, the task of real-time accurate modeling is challenging, mainly because of the complex structure 
of this organ composed of three main constituents: \emph{parenchyma}, \emph{vascular networks} and \emph{Glisson's capsule}.
The parenchyma is certainly the most studied component of the liver; today the researchers agree on hyperelastic 
properties of the tissue, for which the mechanical parameters have been reported for example in~\cite{}. 
Several methods have been proposed to model the hyperelastic behaviour at real-time rates, such multiplicative Jacobian decomposition
introduced in~\cite{stephanie2008}.

@@

The final constituent represented by Glisson's capsule have been already studied in the literature. 
Extensive experiments on porcine liver has been published in~\cite{Umale2013}. The measurements show that although being very 
thin (10--20$\mu m$), the Young's modulus of the capsule (reported as $8.22\pm3.42$\,MPa) exceeds the values for the parenchyma by three orders of magnitude,
suggesting the mechanical influence of the membrane on the liver behaviour.

In~\cite{Hollenstein2006}, a local influence of the capsule has been measured using a special aspiration device. The study was then repeated 
in vivo on human patients during the operation, confirming the mechanical importance of the membrane~\cite{Ahn2010,Nava2008}.
To our best knowledge, on attempt has been made to demonstrate the role of the Glisson's capsule on the global behaviour of 
the liver, i.e. during large deformations of the organ.

The main contribution of the paper is twofold: first, we present an efficient real-time model of the liver, taking into account 
the parenchyma, vascularization and Glisson's capsule. The model is based on three different finite element representations for each constituent,
linked together via mechanical coupling. We show that the model mimics the local experiments descibed in~\cite{Hollenstein2006}.

@@

It is known that the parenchyma exhibits non-linear viscoelastic behaviour \cite{marchesseauMJED}.
However, as we are mainly interested in the static equilibrium, we do not model the time-dependent
phenomena related to viscosity.

@@

The parenchyma is modeled using corotational finite elements~\cite{Felippa}.
While relying on linear stress-strain relationship, it allows for large displacements including rotations. 
While in the full non-linear formulation, the stiffness matrix relates the forces a $\Vec{f}$ and 
displacements $\Vec{u}$ as $\Vec{f} = \Mat{K}(\Vec{u})$, the corotational model 
requires the stiffness matrix $\Mat{K}_0$ of the system be computed only once before the simulation begins. 
Then, in each step, the motion of each element $e$ is decomposed into rigid rotation $\Mat{R}^e$ and local deformation. 
The rotations are then used to update each local element stiffness matrix as $\Mat{R}\Mat{K}_0\Mat{R}^{\top}$
whereas the deformations are used to compute the linear strain in the local corotational frame.
%The stiffness matrix $\Mat{K}$ can be expressed in terms of the deformation
%$u$ and the force as $\Vec{f} = \Mat{K}(\Vec{u}) \Vec{u}$. 
%Where the
%components of $\Mat{K}$ depend on the orientation of each element in the
%current simulation step.
%the rigid component is  
% The corotational
%formulation allows us to deal with geometrical non-linearities but still
%relies on linear stress-strain relationship. Thus rotations and large
%deformations are modelled correctly. 
%In the corotational formulation the deformation
%of each element is expressed in the local frame of reference. 
There are several ways of computing the corotational frame for elements; we rely on
the geometrical method proposed in \cite{Nesme2005}.
% NOTE: This description of corotational method is very simplified and could be extended.

The model of the vascularization is based on linear beams 
with local frames of reference; in many aspects it's similar to the
corotational formulation described above~\cite{Duriez}. As such the model also handles geometric
non-linearities in the deformation. Through a specification of cross section and moments of inertia, 
the model can account for the specific properties of the blood vessels.

The beam-based model of vessels is mechanically coupled to the parenchyma as described in~\cite{peterlik2012}. 
The coupling assumes that there is no relative motion between the vessels and surrounding parenchyma. 
%The proposed model assumes no relative
%motion of vessels to the parenchyma. Segmented vascular network is
%discretized into the set of linear elements that are then projected into
%the tetrahedral mesh of the parenchyma. 
During the simulation, the nodes of linked beams are first displaced and rotated according to the actual motion of associated tetrahedra. 
As the deformation of beams results in mechanical response represented by forces and torques, these are propagated back to 
the tetrahedral FE model.

@@

In the simulations we used material properties for a porcine liver as
reported by Umale \cite{Umale2013}. The
parenchyma was modeled with elastic modulus 1980~Pa and the capsule has
elastic modulus 8~MPa and thickness 20~$\mu$m.
\IP{OK, now we must decide what value we are going to use for the parenchyma, so it does not explode but it's somehow reasonable...)}

@@

\IP{we have to also add some information how we got the models -- either we comment briefly in each subsection:
Local deformations: cube meshed in gmsh (graded mesh?), number of elements, etc...
Global deformations: contrast-enhanced CT data obtained via acquisition performed on a female pig; the 3D mesh created by a
semi-automatic segmentation using ITKSnap~\cite{itksnap}, then meshed using CGAL~\cite{cgal-meshingAlg} resulting in a mesh having 
NNN elements. The capsule modelled using the surface triangular elements extracted from the tetrahedra. The vascularization 
created from the segmentation of hepatic vein as described in~\cite{peterlik2012}. The structured composed of MMM elements.}

@@

\Figure{4in}{aspiration}{TODO: aspiration test; fix figure, add screenshot}

>>>>

@@

In case of human liver, the task of real-time accurate modeling is challenging, mainly because of the complex structure 
of this organ composed of three main constituents: \emph{parenchyma}, \emph{vascular networks} and \emph{Glisson's capsule}.
The parenchyma is certainly the most studied component of the liver; today the researchers agree on hyperelastic 
properties of the tissue, for which the mechanical parameters have been reported for example in~\cite{}\TODO. 
Several methods have been proposed to model the hyperelastic behaviour at real-time rates, such as multiplicative Jacobian decomposition
introduced in~\cite{Marchesseau2010}.

@@

The final constituent represented by Glisson's capsule has been already studied in the literature. 
Extensive experiments on porcine liver have been published in~\cite{Umale2013}. The measurements show that although being very 
thin (10--20$\mu m$), the Young's modulus of the capsule (reported as $8.22\pm3.42$\,MPa) exceeds the values for the parenchyma by three orders of magnitude,
suggesting the mechanical influence of the membrane on the liver behaviour.

In~\cite{Hollenstein2006}, a local influence of the capsule has been measured using a special aspiration device. The study was then repeated 
in vivo on human patients during the operation, confirming the mechanical importance of the membrane~\cite{Ahn2010,Nava2008}.
To our best knowledge, no attempt has been made to demonstrate the role of the Glisson's capsule on the global behaviour of 
the liver, i.e. during large deformations of the organ.

The main contribution of the paper is twofold: first, we present an efficient real-time model of the liver, taking into account 
the parenchyma, vascularization and Glisson's capsule. The model is based on three different finite element representations for each constituent,
linked together via mechanical coupling. We show that the model mimics the local experiments described in~\cite{Hollenstein2006}.

@@

It is known that the parenchyma exhibits non-linear viscoelastic behaviour \cite{Marchesseau2010}.
However, as we are mainly interested in the static equilibrium, we do not model the time-dependent
phenomena related to viscosity.

@@

The parenchyma is modeled using corotational finite elements~\cite{Felippa2005}.
While relying on linear stress-strain relationship, it allows for large displacements including rotations. 
While in the full non-linear formulation, the stiffness matrix relates the forces a $\Vec{f}$ and 
displacements $\Vec{u}$ as $\Vec{f} = \Mat{K}(\Vec{u})$, the corotational model 
requires the stiffness matrix $\Mat{K}_0$ of the system be computed only once before the simulation begins. 
Then, in each step, the motion of each element $e$ is decomposed into rigid rotation $\Mat{R}^e$ and local deformation. 
The rotations are then used to update each local element stiffness matrix as $\Mat{R}\Mat{K}_0\Mat{R}^{\top}$
whereas the deformations are used to compute the linear strain in the local corotational frame.
%The stiffness matrix $\Mat{K}$ can be expressed in terms of the deformation
%$u$ and the force as $\Vec{f} = \Mat{K}(\Vec{u}) \Vec{u}$. 
%Where the
%components of $\Mat{K}$ depend on the orientation of each element in the
%current simulation step.
%the rigid component is  
% The corotational
%formulation allows us to deal with geometrical non-linearities but still
%relies on linear stress-strain relationship. Thus rotations and large
%deformations are modelled correctly. 
%In the corotational formulation the deformation
%of each element is expressed in the local frame of reference. 
There are several ways of computing the corotational frame for elements; we rely on
the geometrical method proposed in \cite{Nesme2005}.
% NOTE: This description of corotational method is very simplified and could be extended.

The model of the vascularization is based on linear beams 
with local frames of reference; in many aspects it's similar to the
corotational formulation described above~\cite{Duriez2006}. As such the model also handles geometric
non-linearities in the deformation. Through a specification of cross section and moments of inertia, 
the model can account for the specific properties of the blood vessels.

The beam-based model of vessels is mechanically coupled to the parenchyma as described in~\cite{Peterlik2012}. 
The coupling assumes that there is no relative motion between the vessels and surrounding parenchyma. 
%The proposed model assumes no relative
%motion of vessels to the parenchyma. Segmented vascular network is
%discretized into the set of linear elements that are then projected into
%the tetrahedral mesh of the parenchyma. 
During the simulation, the nodes of linked beams are first displaced and rotated according to the actual motion of associated tetrahedra. 
As the deformation of beams results in mechanical response represented by forces and torques, these are propagated back to 
the tetrahedral FE model.

@@

In the simulations we used material properties for a porcine liver as
reported by Umale \cite{Umale2013}. The
parenchyma was modeled with elastic modulus 1980~Pa and the capsule has
elastic modulus 8~MPa and thickness 20~$\mu$m.
\IP{OK, now we must decide what value we are going to use for the parenchyma, so it does not explode but it's somehow reasonable...)}
\TG{Right now I'm using 3500 Pa for parenchyma. I can even use higher
modulus, but not any smaller because we get large deformations and use of
corotational model might become questionable.}

@@

\IP{we have to also add some information how we got the models -- either we comment briefly in each subsection:
Local deformations: cube meshed in gmsh (graded mesh?), number of elements, etc...
Global deformations: contrast-enhanced CT data obtained via acquisition performed on a female pig; the 3D mesh created by a
semi-automatic segmentation using ITKSnap~\cite{itksnap}, then meshed using CGAL~\cite{cgal-meshingAlg} resulting in a mesh having 
NNN elements. The capsule modelled using the surface triangular elements extracted from the tetrahedra. The vascularization 
created from the segmentation of hepatic vein as described in~\cite{Peterlik2012}. The structured composed of MMM elements.}
\TG{Sure. Put it here or into the respective subsections? (cube: graded
  mesh created by gmsh with 2648 tetrahedra, 518 triangles for capsule
extracted from tetrahedral mesh.)}

@@

\begin{figure}[htb]
  \centering
  \begin{tabular}{cc}
    \includegraphics[width=3in]{aspiration.pdf} &
    \includegraphics[height=1.5in]{aspiration.jpg}\\
    a) & b)
  \end{tabular}

  \caption{\label{fig-aspiration}Aspiration test: a) displacment profile
  with (blue) and without (black) the capsule; b) simulation scene}
\end{figure}
